Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-13 aistats.tex
Remove excess $ sign
Update abstract and create related work section

\begin{abstract}
We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family.
Surprisingly, we found no general solution to this problem in the literature.
In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD),
yet it falls out of scope for modern analysis.
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

@@

Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
A contrario the KL is not smooth, and it can be infinite in misspecified settings.
But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.

@@

Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
\[
R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
\]
where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$. Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule. Unfortunately however, their rate are worse than the offline MAP or MLE algorithms. For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$. From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$. Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$. This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.

@@

@@

\paragraph{Notation}
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

@@

\subsection{Quadratic Case}
As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
\[ \cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).\]
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}.
	=\frac{\Var(T)}{2n}
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields to
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

@@

\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as
\citet{anastasiou2017bounds},
\citet{marteauferey2019beyond})

@@

>>>>

\begin{abstract}
We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way. 
Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime. 
After exhibiting various facets of the problem, we show the MAP can be interpreted as running stochastic mirror descent (SMD) on the log-likelihood, 
yet modern convergence results do not apply for standard examples of exponential family, highlighting holes in the convergence literature. 
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

@@

@@

@@

\subsection{Related Work}

\paragraph{Density Estimation}
Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
A contrario the KL is not smooth, and it can be infinite in misspecified settings.
But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.

\paragraph{Online Learning}
Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
\[
R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
\]
where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$. Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule. Unfortunately however, their rate are worse than the offline MAP or MLE algorithms. For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$. From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$. Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$. This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.

@@

\subsection{Notation}
$X, T(X)$ and $T$ are random variables, while $x$ is a sample. 
$n$ is the number of samples and $d= \dim(T)$. 
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

@@

\subsection{Quadratic Case}
As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
\[ \cX=\real,\quad \nu(dx) = \exp(\half[-x^2])dx,\quad T(X)=X.\]
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}.
	=\frac{\Var(T)}{2n}
	=\frac{1}{2n}
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields to
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

@@

@@

\TODO{Incorporate this content} ( about \citet{ostrovskii2021finite} as well as
\citet{anastasiou2017bounds} and
\citet{marteauferey2019beyond})
\fdk{
We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate 
%[Robust stochastic approximation approach to stochastic programming]. (The paper uses a bounded stochastic gradient assumption, but can be relaxed to bounded variance instead.) However, strong-convexity in the Euclidean norm is not strictly necessary to find an algorithm with convergence rate of $O(1/n)$, as shown by [Non-strongly-convex smooth stochastic approximation with convergence rate O(1/n)] for linear and logistic regression. Their proof techniques based on self-concordance have been extended to other setting, [Finite-sample analysis of M-estimators using self-concordance, Learning Exponential Families in High-Dimensions: Strong Convexity and Sparsity].
%The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
%The paper [Finite-sample analysis of 𝑀-estimators using self-concordance*] has a good background section on the work of local asymptotic normality (1.1 Classical asymptotic theory) that might be useful. The first 2-3 pages of [PARAMETRIC ESTIMATION. FINITE SAMPLE THEORY] are also nice
}
